Explore and Summarize data (red wine quality) by Nada Alsaab
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages that you end up using in your analysis in this code
library(ggplot2)
library(dplyr)
library(gridExtra)
library(RColorBrewer)
library(GGally)
library(reshape)
library(plotly)
library(rmarkdown)

```

```{r global_options, include=FALSE}

# In HTML format avoid the additional info of the graphs and matrices
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```

```{r echo=FALSE, Load_the_Data}

# Load the Data
wine_data <- read.csv("wineQualityReds.csv")
wine_data$quality <- factor(wine_data$quality, order=TRUE)

```

# Introduction
In this project, Red Wine Quality data is used to know the relationship between the wine features.

## Start data exloratory

```{r echo=FALSE, names}

# Show the columns names
names(wine_data)

```

```{r echo=FALSE, structure}

# Show the data structure
str(wine_data)

```
This dataset has 1599 observations and 13 varibales. 
quality variable has two columns, one is ordinal and other is numeric.

```{r echo=FALSE, summary}

# show the summary of the data
summary(wine_data)

```
Median quality is 5.68, while mean quality is almost 6.64.


# Univariate Plots Section

```{r echo=FALSE, fun}

# Implement a function that drawing the histogram
histogram_fun <- function(var, xlabel)
{
  return(qplot(x=var, data=wine_data, xlab=xlabel))
}

```


```{r echo=FALSE, var1}

# Call the histogram function with data of fixed.acidity
histogram_fun(wine_data$fixed.acidity, 'fixed.acidity (tartaric acid - g / dm^3) ')

```


This is a normal distributed data. The range of fixed acidity is between 4 and 16. The most fixed acidity is between 7 and 8.



```{r echo=FALSE, var2}

# Call the histogram function with data of volatile.acidity
histogram_fun(wine_data$volatile.acidity, 'volatile.acidity (acetic acid - g / dm^3)')

```


This is a normal distributed data. The range of volatile acidity is between 0.1 and 1.6. The most volatile acidity is between 0.4 and 0.7.




```{r echo=FALSE, var3}

# Call the histogram function with data of citric.acid
histogram_fun(wine_data$citric.acid , 'citric.acid (g / dm^3)')

```


This is a right skewed distribution. The range of the critic acid is between 0 and 1. 




```{r echo=FALSE, var4}

# Call the histogram function with data of residual.sugar
histogram_fun(wine_data$residual.sugar, 'residual.sugar (g / dm^3)')

```


This is a right skewed distribution. The range of residual sugar is between 1 and 16. The most residual sugar is between 1.5 and 3.


```{r echo=FALSE, var5}

# Call the histogram function with data of chlorides
histogram_fun(wine_data$chlorides , 'chlorides (sodium chloride - g / dm^3)')
 
```


This is a normal distribution. The range of chlorides is between 0.0 and 0.65. The most chlorides is between 0.05 and 0.1.


```{r echo=FALSE, var6}

# Call the histogram function with data of free.sulfur.dioxide
histogram_fun(wine_data$free.sulfur.dioxide , 'free.sulfur.dioxide (mg / dm^3)')

```


This is a right skewed distribution. The range of free sulfur dioxide is between 0 and 72. The most free sulfur dioxide is between 1 and 10.

```{r echo=FALSE, var7}

# Call the histogram function with data of total.sulfur.dioxide
histogram_fun(wine_data$total.sulfur.dioxide , 'total.sulfur.dioxide (mg / dm^3)')

```


This is a right skewed distribution. The range of the total sulfur dioxide is between 0 and 300. The most total sulfur dioxide is between 10 and 50.


```{r echo=FALSE, var8}

# Call the histogram function with data of density
histogram_fun(wine_data$density , 'density (g / cm^3)')

```


This is a normal distribution. The range of density is between 0.99 and 1.005. The most density is between 0.995 and 1.

```{r echo=FALSE, var9}

# Call the histogram function with data of pH
histogram_fun(wine_data$pH , 'pH')

```


This is a normal distribution. The range of pH is between 2.75 and 4.25. The most pH is between 3.25 and 3.5.


```{r echo=FALSE, var10}

# Call the histogram function with data of sulphates
histogram_fun(wine_data$sulphates , 'sulphates (potassium sulphate - g / dm3)')

```


This is a normal distribution. The range of sulphates 0.25 and 2. The most sulphate is between 0.5 and 0.75.

```{r echo=FALSE, var11}

# Call the histogram function with data of alcohol
histogram_fun(wine_data$alcohol , 'alcohol (% by volume)')

```


This is a right skewed distribution. The range of alcohol is between 8 and 15. The most alcohol is between 9 and 10.


```{r echo=FALSE, var12}

# Call the histogram function with data of quality
histogram_fun(wine_data$quality , 'quality')

```


Quality is categorical, the range is between 3 and 8, the highest value is at 5.

I will recategorize the quality as a rank.
low = < 5 
good = (from 5 to 6)
very good = > 6

```{r, recategorize}

# Recategorize the quality as rank (low, good, very good)
wine_data$rank <- ifelse(wine_data$quality < 5, 'low', ifelse(
  wine_data$quality < 7, 'good', 'very good'))
wine_data$rank <- ordered(wine_data$rank, levels = c('low', 'good', 'very good'))

# Visualize the quality rank
ggplot(data = wine_data, aes(x=rank, fill=rank)) +
  geom_bar() + theme_minimal() +
  scale_fill_brewer(type = 'seq', palette = 4)

```


# Univariate Analysis

### What is the structure of your dataset?
This dataset has 1599 observations and 12 variables. 

### What is/are the main feature(s) of interest in your dataset?
Quiality is the main feature interest. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I think Alcohol, pH, volatile acidity and total sulfur dioxide.

### Did you create any new variables from existing variables in the dataset?
Yes, I created rank variable to recategorize the quality as (low, good, very good).

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The data set is tidy and good and I did not do anything to it. I use it as it is.

#-----------------------------------------------------------------------------

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, Bivariate_Plots, fig.width=12, fig.height=12}

names <- c('fixed.acidity', 'volatile.acidity', 'citric.acid', 'residual.sugar', 'chlorides', 'free.sulfur.dioxide', 'total.sulfur.dioxide', 'density', 'pH', 'sulphates', 'alcohol', 'quality2')
ggscatmat(wine_data[names]) + theme_grey(base_size=10) 

```


The higher absolute value of correlation coeffeceint, the higher the relationship between the two factors.
From the matrix, we can notice that the highest absolute value of correlation coeffeceint is between pH and fixed.acidity with a value of -0.68. The second highest relationship is between density and fixed.acidity with 0.67, and also between citric.acid and fixed.acidity with 0.67.

```{r, boxPlot1}

# Draw a box plot for pH and Fixed acidity
ggplot(aes(x = pH, y = fixed.acidity), data = wine_data) +
geom_point(position = position_jitter(h = 0), color="purple", alpha = 1/10) +
stat_smooth(method = 'lm')+
labs(title="pH VS Fixed acidity", 
       x="pH", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a negative strong relationship between pH and fixed.acidity

# density vs fixed.acidity 
```{r,Mul pboxPlot2}

# Draw a box plot for Density and Fixed acidity
ggplot(aes(x = density, y = fixed.acidity), data = wine_data) +
geom_point(position = position_jitter(h = 0), color="purple", alpha = 1/10) +
stat_smooth(method = 'lm') +
labs(title="Density VS Fixed acidity", 
       x="Density (g / cm^3)", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a positive strong relationship between density and fixed.acidity


```{r, boxPlot3}

# Draw a box plot for Citric acid and Fixed acidity
ggplot(aes(x = citric.acid, y = fixed.acidity), data = wine_data) +
geom_point(position = position_jitter(h = 0), color="purple", alpha = 1/10) +
stat_smooth(method = 'lm') +
labs(title="Citric acid VS Fixed acidity", 
       x="Citric acid (g / dm^3)", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a positive strong relationship between citric.acid and fixed.acidity


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
After the analysis, the most two factors that have the stronest relationship is between pH and fixed.acidity.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
From the matrix, I notices that most factor that affects the quality is "alcohol" with correlation coeffecenit equals 4.8.

### What was the strongest relationship you found?
The strongest relationship between all the factors is between fixed.acidity and citric.acid.

#-----------------------------------------------------------------------------

# Multivariate Plots Section

```{r, boxPlot4}

# Draw a box plot for Quality and Alcohol colorized by rank
ggplot(aes(x = quality, y = alcohol), data = wine_data) +
  geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/3) + geom_boxplot(alpha = 0.5) + scale_colour_brewer(palette=3)+
labs(title="Quality VS Alcohol", 
       x="Quality", y ="Alcohol (% by volume)")

```


We can notice from the figure above that the most distribution is in between 4.5 and 7.5. And there is a positive relationship between quality and alcohol.


```{r, boxPlot5}

# Draw a box plot for Quality and Volatile acidity colorized by rank
ggplot(aes(x = quality, y = volatile.acidity), data = wine_data) +
  geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/3) + geom_boxplot(alpha = 0.5) + scale_colour_brewer(palette=3)+
  labs(title="Quality VS Volatile acidity", 
       x="Quality", y ="Volatile acidity (acetic acid - g / dm^3)")

```


We can notice from the figure above that the most distribution is in between 4.5 and 7.5. And there is a negative relationship between quality and volatile acidity.


```{r, boxPlot6}

# Draw a box plot for pH and Fixed acidity colorized by rank
ggplot(aes(x = pH, y = fixed.acidity), data = wine_data) +
geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/10) +
stat_smooth(method = 'lm')+
labs(title="pH VS Fixed acidity", 
       x="pH", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a negative strong relationship between pH and fixed acidity and the most rank is "good".

```{r, boxPlot7}

# Draw a box plot for Density and Fixed acidity colorized by rank
ggplot(aes(x = density, y = fixed.acidity), data = wine_data) +
geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/10) +
stat_smooth(method = 'lm') +
labs(title="Density VS Fixed acidity", 
       x="Density (g / cm^3)", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a positive strong relationship between density and fixed acidity. Also, we can notice that the "good" rank has the strongest relationship.


```{r, boxPlot8}

# Draw a box plot for Citric acid and Fixed acidity colorized by rank
ggplot(aes(x = citric.acid, y = fixed.acidity), data = wine_data) +
geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/10) +
stat_smooth(method = 'lm') +
labs(title="Citric acid VS Fixed acidity", 
       x="Citric acid (g / dm^3)", y ="Fixed acidity (tartaric acid - g / dm^3)")

```


From the figure above, we can notice that there is a positive strong relationship between citric.acid and fixed acidity. 


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
I found that there is a strong negative relationship between pH and fixed.acidity, it is most strong for good rank. Also, there is a strong positive relationship between density and fixed.acidity and most strong for good rank.

### Were there any interesting or surprising interactions between features?
That alcohol increase the wine quality while volatile acidity decrease it.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r , plot1}

# Visualize the quality rank
ggplot(data = wine_data, aes(x=rank, fill=rank)) +
  geom_bar() + theme_minimal() +
  scale_fill_brewer(type = 'seq', palette = 4)

```

### Description One
Snice quality has 6 categoried which is as a number, I decide to recategorize them to be more clear, and this figure describe the categories. (low, good, very good)

### Plot Two
```{r, plot2}

# Draw a box plot for fixed.acidity and pH
ggplot(aes(x = pH, y = fixed.acidity), data = wine_data) +
geom_point(position = position_jitter(h = 0), color="purple", alpha = 1/10) +
stat_smooth(method = 'lm')
labs(title="pH VS Fixed acidity", 
       x="pH", y ="Fixed acidity (tartaric acid - g / dm^3)")

```

### Description Two
The correlation coefficient between pH and Fixed acidity is negative and that indicates there a reverse relationship. Also, the correlation coefficient has the highest absolute value which indicates that the relationship is the strongest. 


### Plot Three
```{r, plot3}

# Draw a box plot for fixed.acidity and pH colorized by rank
ggplot(aes(x = pH, y = fixed.acidity), data = wine_data) +
geom_point(aes(color = rank, fill = rank), position = position_jitter(h = 0), alpha = 1/10) +
stat_smooth(method = 'lm')+
labs(title="pH VS Fixed acidity", 
       x="pH", y ="Fixed acidity (tartaric acid - g / dm^3)")

```

### Description Three
The correlation coefficient between pH and Fixed acidity is negative and that indicates there a reverse relationship. Also, the correlation coefficient has the highest absolute value which indicates that the relationship is the strongest. And as we notice from the figure above, most red wines are in "good" quality.

# Reflection
First, I chose red wine dataset which has 1599 kind of wines and 12 variables, I start to explore the data and understand it by exploring and visualizing every variable.
Then, I tried to answer the questions by finding their answers from analysis, visualization and matrices. I tried to find the relationship between factors, which is the the strongest relatioship, which is positive and which is negative.
I noticed that the strogest relationship is between pH and Fixed acidity but it is a negative relationship.
I explored the most factors that affect the wine quality, the most effective factor is alcohol, then  volatile acidity.
I enjoyed this analysis and I think IF we have larger dataset we can have more findings.